---
title: "Kansas City R Users Group February"
author: "Jared Harpole"
date: "February 11, 2017"
output: 
      slidy_presentation:
            font_adjustment: 2
          
---

## Manipulating Files Why Do I Need to Know This?
* For interactive work you may not need to do this
      + However, I will show you a cool trick to select files for loading into R
* If you want to automate work flows
      + I am a Data Scientist and I use this
      + Researchers might use as well
* There are built-in functions in R that are probably better than what you could write
      + Allows for cross portability of code across OSs
      + Succint function calls
* This is a Beginners Meetup
      + Some material/approaches are advanced
      + Please stop me if you have questions


## Today's Agenda

1. A quick and easy way to get a file path
2. How to grab the file name from a file path
3. Creating file paths that are operating system independent
4. What is the `~` and how to use it in file paths 
5. How to get important file meta data
6. Testing if file or directories exist and creating them if needed

## A quick and easy way to get a file path

* Have you ever been frustrated by reading data into R?
    + Getting file paths can suck
  
* There is a quick and easy way while you are learning
    + `file.choose`

* Let's Take a Quick Example
    
```{r basen, comment=NA, size="footnotesize", eval=FALSE}
my_file <- file.choose()
my_csv <- read.table(file = my_file, sep = ",", header = TRUE)
```

## How to grab the file name from a file path

* How do you grab the file name from "/home/jared/test.csv" 

* You can write your own function like so

```{r, comment = NA}
## Write a function to split the path (returns a list)
my_file_path <- "/home/jared/test.csv"
splitted_path <- strsplit(x = my_file_path, split = "/")
splitted_path
## Need to flatten the list or grab the 1 element (either 1 works)
# Using unlist to make a list a vector
flatten1 <- unlist(splitted_path)
flatten1
# Using [[]] operator
flatten2 <- splitted_path[[1]]
flatten2
## Now grab the last element (tail or lenght)
# tail version
file_name1 <- tail(x = flatten1, n = 1)
file_name1

file_name2 <- flatten1[length(flatten1)]
file_name2
## Here it is in 1 line
file_name <- tail(unlist(strsplit(x = my_file_path, split = "/")), n = 1)
file_name 
```

## How to grab the file name from a file path Cont....

* What are some problems with the previous approaches?


## How to grab the file name from a file path Cont...

* There is a better way (Thank You R Base!!!)

* R has a function called basename

```{r, comment = NA}
my_file_path <- "/home/jared/test.csv"

file_name <- basename(path = my_file_path)
file_name
```

## Creating file paths that are operating system independent

* Can't you just use `paste`? Yes but there are problems

```{r, comment = NA}
## Avoid doing this
my_bad_linux_path <- paste(c("/home", "jared", "jared.csv"), collapse = "/")
my_bad_linux_path

my_bad_windows_path <- paste(c("C:", "User", "Jared"), collapse = "\\")
my_bad_windows_path
```

* If You go from Windows to Linux you will have problems potentially

* A better option is file.path

```{r, comment = NA}
my_good_linux_path <- file.path("/home", "jared", "jared.csv")
my_good_linux_path

my_good_windows_path <- file.path("C:", "User", "Jared", "jared.csv")
my_good_windows_path
```


## What is the `~` and how to use it in file paths 

* The `~` operator usually does /home/<username>
      + /home/jared
      + /home/ubuntu
        
* Typically this is nice because then you don't have to write out /home/<username>
      + Extremely useful for making software work depending on user executing
      + Recommender Server on AWS 

* In R you can use path.expand to take advantage of `~`
```{r, comment = NA}
## Let's use file.path and path.expand
file_path <- file.path("~", "jared.csv")
print(file_path)
path.expand(file_path)
```

## How to get important file meta data

* Let's say I want to know information about a file 
      + The size in bytes
      + The file modification time
      + The last status change

* How do we do this in R? file.info to the rescue

```{r, comment = NA}
test_file <- file.path("~", "FileManip", "test.txt")
test_file <- path.expand(test_file)

file.info(test_file)

## Lets get the size and modification time
file.info(test_file)[c("size", "mtime")]
```

## Testing if file or directories exist and creating them if needed

* Let's say you have a program that you want to test if a file exists
      + If it exists don't do anything and if not create
      + file.exists and file.create to the rescue
      
```{r, comment = NA, echo = FALSE, message = FALSE}
test_file <- path.expand(file.path("~", "kcr.csv"))
if(file.exists(test_file)){
  removed <-file.remove(test_file)
}
```


```{r, comment = NA}
test_file <- file.path("~", "kcr.csv")
test_file <- path.expand(test_file)
## file.exists function
file.exists(test_file)
## Create a boolean variables
exists <- file.exists(test_file)
## Test for existence
if(!exists){
  print("File does not exist. Creating test_file file")
  file.create(test_file)
  print(paste0("Was test_file created: ", file.exists(test_file)))
}
```


## Testing if file or directories exist and creating them if needed Cont..

* Now let's do the same thing for a directory
      + If it exists don't do anything and if not create
      + dir.exists and dir.create to the rescue
      
```{r, comment = NA, echo = FALSE, message = FALSE}
test_dir <- path.expand(file.path("~", "kcr_rulz"))
if(dir.exists(test_dir)){
  removed <- unlink(test_dir)
}
```

```{r, comment = NA}
test_dir <- path.expand(file.path("~", "kcr_rulz"))
test_dir
## Create existence variable with dir.exists
exists <- dir.exists(test_dir)
if(!exists){
  print("Directory does not exist. Creating directory")
  dir.create(test_dir)
  print(paste("Was test_dir created: ", dir.exists(test_dir)))
}
```
