---
title: "Introduction to Recoding Variables in R"
author: "Jared Harpole"
date: "Saturday December 10, 2016"
output: 
      slidy_presentation:
            font_adjustment: 4
          
---

## Why Recode Variables in R?
* We don't have to use R for recoding variables
      + We could use Excel or another program with a GUI
* Using a program with a GUI that involves pointing and clicking is not ideal
      + Lack of reproducibility
      + Difficult and/or impossible to retrace
      + Does not work in a team environment
* By using R and saving our .R files we have a log of what we did
      + This allows for better reproducibility
      + Retracing is tractable
      + Works in team environments


## Today's Agenda

1. Recoding numerics, factors, and characters
2. Using Base R for these
3. Using mapvalues in plyr 
4. Applications to missing data
5. Advanced topic: Using functional programming for recoding values

## Recoding Numerical Data: Base R

```{r basen, comment=NA, size="footnotesize"}
set.seed(1)
dat_1a <- data.frame(a = sample(3, 10, replace = TRUE), b = sample(2, 10, replace = TRUE))
dat_1a
dat_1a$a[dat_1a$a == 1] <- 50
dat_1a$b[dat_1a$b == 2] <- 100
dat_1a
```

* With base R using more complex boolean expressions are possible
* Select values between -0.7 and 0.2 and code as 100
```{r, comment = NA}
set.seed(1)
dat_1b <- data.frame(a = rnorm(10), b = rnorm(10))
dat_1b
dat_1b$a[dat_1b$a < 0.2 & dat_1b$a > -.7] <- 100
dat_1b
```

## Recoding Factors: Base R Take 1
* Using `dat_2$a[dat_2$a == 1] <- 50` does not work for factors
```{r basefw, comment = NA}
dat_2 <- data.frame(a = gl(2, 5, labels = c('hi', 'low')), b = gl(5, 2, labels = letters[1:5]))
dat_2
dat_2$a[dat_2$a == 'hi'] <- 'v-high'
dat_2
```
```{r, echo = FALSE, cache=TRUE}
dat_2a <- data.frame(a = gl(2, 5, labels = c('hi', 'low')), b = gl(5, 2, labels = letters[1:5]))
dat_2b <- dat_2a
```
## Recoding Factors: Base R Take 2
* Okay let's do it right using `levels()`
```{r, comment = NA}
dat_2a
levels(dat_2a$a) <- c('walrus', 'snake')
dat_2a
```
* Note that the order that walrus = hi and snake = low as per original coding




* Quick and dirty reversal of levels with `rev()` on column b
```{r basefr, comment = NA}
dat_2b
levels(dat_2b$b) <- rev(levels(dat_2b$b))
dat_2b
```

## Recoding Characters: Base R 
* Here we can go back to the `dat_3$a[dat_3$a == "a"] <- 'z'` syntax
```{r basec, comment = NA}
set.seed(1)
dat_3 <- data.frame(a = sample(letters[1:3], 10, replace = TRUE), 
                    b = sample(letters[11:14], 10, replace = TRUE), 
                    stringsAsFactors = FALSE)
dat_3$a[dat_3$a == 'a'] <- 'z'
dat_3
```

## Recoding a Continuous Variable as Categorical: Base R
* We will use the `cut()` function to do this
* Note intervals are closed on the right and opened on left by default
* I used ordered_result = TRUE to let R know this is an ordered factor
```{r, comment = NA}
set.seed(1)
dat_4a <- data.frame(a = rnorm(20), b = rnorm(20))
dat_4b <- dat_4a
dat_4a
dat_4a$a <- cut(dat_4a$a, breaks = c(-Inf, -1, 1, Inf), 
                ordered_result = TRUE)
dat_4a$b <- cut(dat_4a$b, breaks = c(-Inf, 0, Inf),
                ordered_result = TRUE)
dat_4a

```
* What if you don't like those labels
```{r, comment = NA}
dat_4b$a <- cut(dat_4b$a, breaks = c(-Inf, -1, 1, Inf), 
                labels = c("small", "med", "large"),
                ordered_result = TRUE)
dat_4b$b <- cut(dat_4b$b, breaks = c(-Inf, 0, Inf), 
                labels = c("low", "high"), 
                ordered_result = TRUE)
dat_4b
```

## Using the mapvalues Function in plyr: Numeric
* Up until know everything we did was using base R functions
* The mapvalues function may now be your best friend
```{r, comment = NA, echo = FALSE}
if(!require(plyr))
      install.packages('plyr')
set.seed(1)
```
```{r, comment = NA}
dat_5 <- data.frame(a = sample(3, 10, replace = TRUE), b = sample(2, 10, replace = TRUE))
dat_5
dat_5$a <- mapvalues(dat_5$a, from = c(1,2,3), to = c(3,2,1))
dat_5
dat_5$b <- mapvalues(dat_5$b, from = c(1,2), to = c(10, 20))
dat_5
```

## Using the mapvalues Function in plyr: Factors
* Here I show how to do a general recode of factors
```{r, comment = NA, cache=TRUE}
dat_6 <- data.frame(a = gl(2, 5, labels = c('hi', 'low')), b = gl(5, 2, labels = letters[1:5]))
dat_6
dat_6$a <- mapvalues(dat_6$a, from = c('hi', 'low'), to = c(1, 2))
dat_6
```

* Here I show how to combine factor levels with mapvalues
* This is more straightforward using mapvalues that Base R (in my opinion)
```{r, comment = NA}
dat_6$a <- mapvalues(dat_6$a, from = c(1, 2), to = c(1,1))
dat_6$b <- mapvalues(dat_6$b, from = letters[1:5], 
                     to = c('a', 'a', 'c', 'd', 'e'))
dat_6
```
* Note that recoding characters works the same way as with numerics:
  + Add the thing to recode `dat_6$a`
  + Add the `from = c('did not participate', 'not applicable')`
  + Add the `to = c(NA, NA)`
  + Run the function and get your recoded values
  
## Pros and Cons of Base R vs mapvalues
* Base R may take slightly less typing for simple cases vs mapvalues
* mapvalues may be more intuitive to use (you be the judge of that)
* Base R works better if you need complex boolean recoding
* mapvalues is more straightforward for combining factor levels
* mapvalues is more straightforward for reverse coding
* Overall I think the mapvalues function is a good option especially for beginners on most tasks

## Missing Data Example with Base R and mapvalues
* Let's look at a case of 100 observations with 5 variables that have 1 missing value
      + The missing value is -999
```{r, cache = TRUE, comment = NA, echo = FALSE}
makeMiss <- function(Nit, Ncase, missVec){
    set.seed(1234)
    sampVals <- c(1:5, missVec)
    dat <- list()
    for(i in 1:Nit){
        dat[[i]] <- sample(sampVals, Ncase, replace = TRUE)
    }
    dat <- as.data.frame(do.call(cbind, dat))
    dat
}
dat_7a <- makeMiss(5, 100, -999)
dat_7b <- dat_7a
dat_8a <- makeMiss(15, 100, -999)
dat_8b <- makeMiss(15, 100, -999)

```

```{r, comment = NA}
head(dat_7a)
sapply(dat_7a, table)
dat_7a$V1[dat_7a$V1 == -999] <- NA
dat_7a$V2[dat_7a$V2 == -999] <- NA
dat_7a$V3[dat_7a$V3 == -999] <- NA
dat_7a$V4[dat_7a$V4 == -999] <- NA
dat_7a$V5[dat_7a$V5 == -999] <- NA
sapply(dat_7a, table, useNA = "always")

```

* Now let's do the mapvalues way
```{r, comment = NA}
head(dat_7b)
sapply(dat_7b, table)
dat_7b$V1 <- mapvalues(dat_7b$V1, from = -999, to = NA)
dat_7b$V2 <- mapvalues(dat_7b$V2, from = -999, to = NA)
dat_7b$V3 <- mapvalues(dat_7b$V3, from = -999, to = NA)
dat_7b$V4 <- mapvalues(dat_7b$V4, from = -999, to = NA)
dat_7b$V5 <- mapvalues(dat_7b$V5, from = -999, to = NA)
sapply(dat_7b, table, useNA = "always")

```

* This approach works okay but there is a serious flaw in using both mapvalues
and Base R
      + What is the serious flaw?
      + Hint what if I have 20 or 100 variables
      
## Advanced Topic: Using Functions For Recoding Missing Values
* Up until this point I showed you how to recode values one at a time
      + This may be a viable way depending on your problem
* I want you to start to be able to see patterns that may make your life easier
* Repeating a task = opportunity to simplify code
* This is where using functions comes in handy 
* Let's take an example

## Advanced Topic Example: 15 variables with one missing value
* Let me show you how we might tackle a problem like this.
```{r, comment = NA}
head(dat_8a)
sapply(dat_8a, table)
```
* Our missing value is -999
* To solve this for 1 variable at a time we could do the following
```{r, comment = NA}
table(dat_8a$V1)
dat_8a$V1[dat_8a$V1 == -999] <- NA
table(dat_8a$V1)
```
* Now if we need to recode 15 variables that may have  missing values how can we do this more efficiently?
* Using a function and applying this function over the data frame
```{r, comment = NA, cache = TRUE}
recodeSingleMiss <- function(x, naVal){
    x[x == naVal] <- NA
    x
}
```
* Notice that `x[x == naVal] <- NA` is `dat_8a$V1[dat_8a$V1 == -999] <- NA`
* Now using this function we can recode 15 variables super fast and don't risk making near the typos
```{r, comment = NA}
head(dat_8b)
sapply(dat_8b, table)
dat_8b[] <- lapply(dat_8b, recodeSingleMiss, -999)
sapply(dat_8b, table, useNA = "always")

```

## Questions and/or Comments 