---
title: 'KC R Users: Subsetting'
author: "Jared Harpole"
date: "April 16, 2016"
output: slidy_presentation
---

# Overview

* Discuss `[` operator with vectors, matrices, data frames and lists
* Discuss the `[[` and `$` operators for data frames and lists
* Discuss simplifying and preserving operations on data structures
* Discuss how to access S3 and S4 objects
* Using `subset` function
* Using lookup tables and random samples of rows
* Using `grep` and `setdiff` for selecting column names (if time permits)
* Using `which` for finding integer locations ????
* Boolean algebra vs. set operations????

# The `[` Operator with Vectors

 * Using integer subsetting

```{r, cache = TRUE, comment=NA, warning=FALSE, message=FALSE}
### vectors
x <- c(1.1, 2.3, 4.5, 1.05, 3.9)
# using [
x[1]
x[c(2,3)]
x[order(x)]
order(x)
```

* Using integer subsetting with `-`

    + Cannot use `x[c(-1, 2)]`
    
```{r, cache = FALSE, comment=NA, warning=FALSE, message=FALSE}
x
### using [ with -
x[-c(1, 4)]
x[-1]


```

* Using integer subsetting with `-`

    + A trick for multiple indices using `:` or `seq`
    
```{r, cache = FALSE, comment=NA, warning=FALSE, message=FALSE}
## Here is a trick for multiple indices
x[-c(1:3)]


set.seed(1)
dog <- rnorm(20)
dog
dog[-c(11:20)]

dog[-c(seq(2, 20, 2))]  
```

* Using boolean subsetting
 

```{r, cache = FALSE, comment=NA, warning=FALSE, message=FALSE}
x
# using [ with booleans
x[c(TRUE, TRUE, FALSE, TRUE, FALSE)]
x[x > 2.5]
x > 2.5 

x[x > 2.5 | x < 1.10]

x > 2.5 | x < 1.10
```

* Using character subsetting

     + Cannot use `y[-c('a', 'b')]` need to use `grep`, `setdiff`, or `subset`
     + More on this later if we have time 
```{r, cache = FALSE, comment=NA, warning=FALSE, message=FALSE}
# using [  with characters
(y <- setNames(x, letters[1:5]))
y
y['a']
y[c('c', 'a', 'd')]

```
   
# The `[` Operator with Matrices

* Using integer subsetting simple cases

```{r, cache = TRUE, comment=NA, warning=FALSE, message=FALSE}
## Matrices
test <- matrix(1:16, nrow = 4)
test

## Simple integer subsetting 
test[1:2, ]
test[, 3:4]
test[c(1, 3), c(1, 2)]
test[1, ]
```

* Using integer subsetting with `-`

```{r, cache = FALSE, comment=NA, warning=FALSE, message=FALSE}
## Can also use -
test

test[-c(1,3), -c(1,2)]
```

* Using integer subsetting with another matrix 

```{r, cache = TRUE, comment=NA, warning=FALSE, message=FALSE}
select <- matrix(c(1, 1, 1, 3, 2, 4), ncol = 2, byrow = TRUE)
select 
test 
test[select]
```

* Using character and boolean subsetting

```{r, cache=FALSE, comment=NA, warning=FALSE, message=FALSE}
dimnames(test) <- list(c('r1', 'r2', 'r3', 'r4'), 
                       c('c1', 'c2', 'c3', 'c4'))
test
test[, 'c1']

test[test[, 'c1'] > 2 & test[, 'c2'] < 10, ]
```

* Matrices are Vectors in Column Major Form

```{r, cache=TRUE, comment=NA, warning=FALSE, message=FALSE}
#vectors behind the scenes in column major form
test
test[1,2]
test[5]
as.vector(test)
```

# The `[` Operator on Data Frames

* Subsetting like a list

```{r, cache=TRUE, comment=NA, warning=FALSE, message=FALSE}
## Data frames
df <- data.frame(a = 1:6, b = 6:1, c = LETTERS[1:6], stringsAsFactors = FALSE)
df
## Like a list
df[c('a', 'b')]
df[1:2]
```

* Subsetting like a matrix

```{r, cache=TRUE, comment=NA, warning=FALSE, message=FALSE}
df
## Like a matrix
df[, c('a', 'b')]
df[, 1:2]
```

* What is the Difference Jared?

```{r, cache=TRUE, comment=NA, warning=FALSE, message=FALSE}
df
## What is the difference?

## Preserving
df['a']

## Simplifying
df[, 'a']
```

* Note using the `-` operator also works with integers in the same way as with a matrix
    + Just be sure to watch out for simplifying and preserving behavior

* Subsettting a data frame using booleans

```{r, comment=NA}
df 
## Subseting a data frame based on column values
df[df['a'] > 2 & df['b'] > 2, ]

## Subsetting a 
df[df['a'] > 2 & df['b'] > 2, c('a', 'c')]
```

# The `[[` and `$` Operators for Lists and Data Frames

* `[[` and `$` are used with data frames and lists to pull pieces out of a list
* Think of the list x as a set of 3 cars on a train (Wickman, 2013, p. 40)

```{r, cache = TRUE, comment=NA}
train <- list(a = c(1, 2, 3), b = c(3, 2, 1), c = letters[1:3])
train
```

* The `[` operator select cars on the train
    + Below selects the first car and both the first and second cars, respectfully

```{r, comment=NA}
## car 1
train[1]

## first 2 cars
train[1:2]
```

* The `[[` or `$` operator select the contents in a given car

    + Select the contents of car 1
    
```{r, comment=NA} 
train[[1]]
```

* Note if you pass a vector to `[[` operator it indexes recursively

```{r, comment=NA}
new_list = list(z = list(y = list(x = list(w = 10))))
new_list

## This 
new_list[[c('z', 'y', 'x', 'w')]]

## And This are the same
new_list[['z']][['y']][['x']][['w']]
```

* `new_list$z` is equivalent to `new_list[['z', exact = FALSE]]`

    + Below `new_list[['z']]` won't work as it will throw and error and not compile
    
```{r, comment=NA}
new_list = list(abc = c(1, 2, 3), def = letters[1:3])
new_list[['a', exact = FALSE]]
new_list$a
```

# Discuss Simplifying and Preserving Operations

* Simplifying subsets return the simpliest data structure
* Preserving subsets return the same structure as the input (ouput = input)
    + This has consequences in programs you use and/or write
    + Wickham (2015, pp. 41-42)
    
|  Structure  |   Simplifying    |              Preserving              |
|:-----------:|:----------------:|:------------------------------------:|
|   Vector    |      x[[1]]      |                 x[1]                 |
|    List     |      x[[1]]      |                 x[1]                 |
|   Factor    | x[1:4, drop = T] |                x[1:4]                |
|   Matrix    | x[1, ] or x[, 1] | x[1, , drop = F] or x[, 1, drop = F] |
| Data Frame  | x[1, ] or x[, 1] |       x[, 1, drop = F] or x[1]       |    

* Vectors

```{r, comment=NA}
x <- c(a = 1, b = 2, c = 3)
#Preserves names
x[1]

#Drops names
x[[1]]
```

* List

```{r, comment=NA}
x <- list(a = c(1, 2, 3), b = letters[5:7])
#Preserves the list object
str(x[2])

#Simplifies list to character vector
str(x[[2]])

```

* Factor

```{r, comment=NA}
fac <- factor(c("A", "B"))
#Preserves all factor levels
fac[1]

#Drops unused levels
fac[1, drop = TRUE]
```

* Matrix

```{r, comment=NA}
mat <- matrix(rnorm(36), ncol = 6)
#Preserves the matrix
mat[, 1, drop = FALSE]

#Drops the matrix to a vector 
mat[, 1]
```

* Data Frame

```{r, comment=NA}
df <- data.frame(a = 1:3, b = 3:1, c = letters[6:8], stringsAsFactors = FALSE)
#Preserves the data frame
df['c']

#Drops the data frame to a character vector
df[['c']]
```

# Accessing S3 and S4 Objects

* S3 Objects are easy we just use `[`, `[[`, or `$`

```{r, comment=NA}
## Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
lm.D9 <- lm(weight ~ group)
class(lm.D9)
str(lm.D9[1:5])
lm.D9['coefficients']
lm.D9[['coefficients']]
lm.D9$coefficients
```

* S4 Objects are a little different

```{r, comment=NA, message = FALSE}
library(Matrix)
mat_ex <- Matrix(1:9, nrow=3,
                dimnames = list(c("a", "b", "c"), c("A", "B", "C")))
str(mat_ex)
## Like $ operator
mat_ex@x
## Like [[ operator
slot(mat_ex, 'x')
```

# The `subset` Function

* The `subset` function will work on vectors, matrices, and data frame
    + I will only cover data frames here but the logic can extend to matrices

* According to the documentation here is how subset for data frames looks:
    + `subset(x, subset, select, drop = FALSE, ...)`

```{r, comment=NA}
df <- data.frame(a = 1:6, b = 6:1, c = letters[6:11], 
                 d = LETTERS[6:11], stringsAsFactors = FALSE)
df
subset(x = df, subset = a > 3 & c %in% c('j', 'k'), select = names(df))

subset(x = df, subset = a > 3 & c %in% c('j', 'k'))
```

# Creating a Lookup Table with Character Subsetting and Random Samples (Bootstrapping)

* Using character matching we can provide a powerful way to create lookup tables

```{r, comment = NA}
## Create data
x <- sample(c('j', 's', 'u'), 15, replace = TRUE)
x
## Create a lookup table
lookup <- c(j = 'Jared', s = 'Steve', u = 'Alien')

lookup[x]

```

* Using integer subsetting to draw random samples or shuffle a data set

  + Here is how you can create a training and testing set from scratch
  
```{r, comment = NA}

df <- data.frame(a = rnorm(15), b = letters[1:15], c = rep(c(T, F, T), times = 5 ), 
stringsAsFactors = FALSE)  

## Without replacement
df[sample(nrow(df), 5, replace = FALSE), , drop = FALSE]
  
  
## With replacement (Bootstrapping)
df[sample(nrow(df), 5, replace = TRUE), , drop = FALSE]
```

* Create a training and testing set

```{r, comment = NA}
df <- data.frame(a = rnorm(15), b = letters[1:15], c = rep(c(T, F, T), times = 5 ), 
stringsAsFactors = FALSE)

train <- sample(nrow(df), floor(nrow(df)*.6), replace = FALSE)

train_df <- df[train, ]

test_df <- df[-train, ]

train_df

test_df
```

# Using `grep` and `setdiff` for Extracting Columns quickly

* `grep` is the global regular expression print 
    + It uses regexs to match patterns and returns 
    + `grep(pattern, x, ...other stuff that I use defaults with)`
* Remember when I said you can't use `-` for negating a character

    + `grep` to the rescue
    
```{r, comment=NA}
df <- data.frame(apple = 1:6, bubble = 6:1, cushion = letters[1:6], dog = LETTERS[1:6], ernie = state.abb[1:6])
df

## We want to remove apple and ernie
purge <- grep("apple|ernie", names(df))
purge
df[, -purge]

```

* A safer way to match with regexs when we have the full names is using `^` and `$` in the string
    + `grep("^apple$|^ernie$", names(df))` 
    + This ensures that the words apple and ernie are matched exactly

* `setdiff` is one of my favorite data manipulation functions

    + It uses set theory to remove elements from the set that do not belong
    + If you know the columns you want to eliminate you can use `setdiff`
    
```{r, comment=NA}
df <- data.frame(apple = 1:6, bubble = 6:1, cushion = letters[1:6], dog = LETTERS[1:6], ernie = state.abb[1:6])
df

keepers <- setdiff(names(df), c('apple', 'ernie'))
keepers
df[, keepers]

```

#Using the `which` function 

* `which` allows you to convert booleans (`TRUE` or `FALSE`) to integers

* Let's take an example:

```{r, comment = NA}
set.seed(12)
ex <- sample(letters, 20, replace = TRUE)
ex
which(ex == 'a')
ex[which(ex == 'a')]
which(ex %in% c('a', 'b', 'c'))

```

#Boolean Algebra vs. Sets

* There is a relationship between logical vectors and their integer equivalents.

* Here we will take several examples:
  
  + `intersect`, `union`, and `setdiff`:
    
    
```{r, comment = NA, cache = TRUE}

(t1 <- 11:20 %% 2 == 0)

(t2 <- which(t1))

(u1 <- 11:20 %% 3 == 0)

(u2 <- which(u1))
```

* `intersect`:

```{r, comment = NA}
# t & u <=> intersect(t, u)
t1 & u1

intersect(t2, u2)
cbind(t1, u1)

```

* `union`:

```{r, comment = NA}
# t | u <=> union(t, u)
t1 | u1

union(t2, u2)
cbind(t1, u1)
```

* `setdiff`:

```{r, comment = NA}
#t & !u <=> setdiff(t, u)
t1 & !u1

setdiff(t2, u2)
cbind(t1, u1)
```