---
title: 'KC R Users: Apply Function Intro'
author: "Jared Harpole"
date: "April 15, 2016"
output: slidy_presentation
---

#Overview

* What are the `apply` family of functionals

* Why use the `apply` family of functionals?

* What exactly are the `apply` family of functionals doing under the hood?

* Review of `lapply`, `sapply`, `vapply`, `apply`, `tapply`, and `mapply`


#What are the `apply` family of functionals and Why use them?

* The `apply` family are a set of functions that allow for writing compact code by 
avoiding loops. 
  + More specifically `apply` and variants are functionals
  + This means that they take a function as input and return a vector as output
  
* What if you want to compute the column sums of a matrix? How would you do it?

  + First way `for` loop:
  
```{r, comment = NA}
set.seed(12)
x <- matrix(rnorm(100), ncol = 10)

output <- vector('numeric', length = ncol(x))
for(col in seq_along(x[1, ])){
  output[col] <- sum(x[,col])
}
output
```

  + Second way `apply`:
  
```{r, comment = NA}
set.seed(12)
x <- matrix(rnorm(100), ncol = 10)

apply(x, MARGIN = 2, FUN = sum)

```

* One line versus 4 lines....You decide?


#Using `lapply` to Illustrate What is Going On?

* Previously we looked at how `apply` worked on a matrix.

* Now I will use `lapply` to give you some intuition of what is going on.

* Let's take another example with lists computing the column sums of each matrix in a list:

  + `for` loop way:
  
```{r, comment = NA}
set.seed(12)
mat_list <- list(a = matrix(rnorm(100), ncol = 10), b = matrix(rlnorm(100), ncol = 10))

out_list <- vector('list', length = length(mat_list))
for(i in seq_along(mat_list)){
  tmp <- vector('numeric', length = ncol(mat_list[[i]]))
  for(j in seq_along(mat_list[[i]][1, ])){
    tmp[j] <- sum(mat_list[[i]][, j])
  }
  out_list[[i]] <- tmp
}
out_list
```

  + `lapply` way:
  
```{r, comment = NA}
set.seed(12)
mat_list <- list(a = matrix(rnorm(100), ncol = 10), b = matrix(rlnorm(100), ncol = 10))

lapply(mat_list, FUN = colSums, na.rm = TRUE)

```

* In this example it made a huge difference

#`lapply` Under the Hood

* In the previous example `lapply` was much more compact and expressive as to what we were doing

* According to the R docs here is how to use `lapply`:
  
    + `lapply(X, FUN, ...)`
  
    + `X` is a vector and will be coerced as a list
  
    + `FUN` is the function to apply to each element of `X`
  
    + `...` are optional arguments to `FUN`
  
* What is `lapply` doing? Let's write our own simple R implementation.


```{r, comment = NA, cache = TRUE}

lapply_jh <- function(X, FUN, ...){
  output <- vector('list', length(X))
  for(i in seq_along(X)){
    output[[i]] <- FUN(X[[i]], ...)  
  }
  output
}

```

* We can see `lapply_jh` is a wrapper for a common `for` loop pattern:

    + Create a container for output `output`
    + Apply a function `FUN` to each component of the list
    + Fill `output` with the results
    
* The logic here generalizes to other members of the `apply` family with various nuances

* Now watch this!!!

```{r, comment = NA}
set.seed(12)
mat_list <- list(a = matrix(rnorm(100), ncol = 10), b = matrix(rlnorm(100), ncol = 10))

lapply(mat_list, colSums, na.rm = TRUE)
lapply_jh(mat_list, colSums, na.rm = TRUE)
```

#Friends of `lapply`: `sapply` and `vapply`

* `sapply` and `vapply` are similary to `lapply` except that they simplify the output if possible

    + `sapply` guesses and `vapply` takes an additional argument to specify the output type
    
* Let's take an example where output can be simplified:

```{r, comment = NA}
set.seed(12)
test_list <- list(x = matrix(rnorm(100), ncol = 10), y = matrix(rnorm(100), ncol = 10))

lapply(test_list, colMeans)

sapply(test_list, colMeans)

vapply(test_list, colMeans, vector('numeric', length = 10))


```

* Let's take an example where output should not be simplified (should throw error):

```{r, comment = NA}
set.seed(12)
test_list <- list(x = matrix(rnorm(100), ncol = 10), y = matrix(rnorm(100), ncol = 5))

sapply(test_list, colMeans)

#vapply(test_list, colMeans, vector('numeric', length = 10))

```

#`tapply`: Grouped Apply
* `tapply` is like a generalization of `apply` to summarizing grouped data

* let's say we have ages from 3 groups that we want to compute the mean age per group:

```{r, comment = NA, cache = TRUE}
ages <- sample(18:65, 45, replace = TRUE)
group <- gl(n = 3, k = 15, labels = c('A', 'B', 'C'))

tapply(X = ages, INDEX = group, FUN = mean, na.rm = TRUE, simplify = TRUE)
```

* We can create a simple R implementation of `tapply`

```{r, comment = NA}
tapply_jh <- function(X, INDEX, FUN, ..., simplify = TRUE){
  groups <- split(X, INDEX)
  sapply(groups, FUN, ..., simplify = TRUE)
}

```

  + First split the data using `split`:
  
```{r, comment = NA}
split(ages, group)
```

  + Now apply `FUN` to groups using `sapply` with simpligy is True
  
```{r, comment = NA}
groups <- split(ages, group)
sapply(groups, FUN = mean, na.rm = TRUE)
```